The \texttt{mc2d} package.

### 2010-02-03 pkg/mc2d/inst/doc/docmcEnglish.rnw
Test for a Vignette (once again)
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/riskassessment@84 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

%% LyX 1.6.2 created this file.  For more info, see http://www.lyx.org/.
%% Do not edit unless you really know what you are doing.
\documentclass[10pt,english]{article}
\usepackage[T1]{fontenc}
\usepackage[latin9]{inputenc}
\usepackage[letterpaper]{geometry}
\geometry{verbose,tmargin=2.5cm,bmargin=2.5cm,lmargin=2.5cm,rmargin=2.5cm}
\setlength{\parskip}{\medskipamount}
\setlength{\parindent}{0pt}
\usepackage{url}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{setspace}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% LyX specific LaTeX commands.
%% Because html converters don't know tabularnewline
\providecommand{\tabularnewline}{\\}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% Textclass specific LaTeX commands.
\usepackage{noweb}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% User specified LaTeX commands.
% sweave commands for vignette
%\VignetteIndexEntry{Tools for Two-Dimensional Monte-Carlo Simulations}
%\VignettePackage{mc2d}

\usepackage{babel}

\begin{document}

\title{The {}``\texttt{mc2d}'' package. }

\author{R. POUILLOT, M.-L. DELIGNETTE-MULLER, D.L. KELLY \& J.-B. DENIS}

\maketitle
\tableofcontents{}

\noindent \begin{center}
\vfill{}

\par\end{center}

This documentation is intended for readers with:
\begin{itemize}
\item A medium level of experience in R. Please refer to the Manual {}``An
Introduction to R'' available with R distribution if needed;
\item Some knowledge about Monte-Carlo simulation (its basic principles
and its utility) and about Quantitative Risk Assessment (QRA). 
\end{itemize}
This documentation will not describe all arguments of the functions.
The definitive reference remains the documentation associated with
the package.

\section{Introduction}

\subsection{What is \texttt{mc2d}?}

{}``\texttt{mc2d}'' means Two-Dimensional Monte-Carlo (\emph{{}``Monte-Carlo
à Deux Dimensions''}). This package :
\begin{itemize}
\item provides additional probability distributions;
\item provides tools to construct One-Dimensional and Two-Dimensional Monte-Carlo
Simulations;
\item provides tools to analyse One-Dimensional and Two-Dimensional Monte-Carlo
Simulations.
\end{itemize}
In a previous version, some tools to fit parametric distributions
to data were included. Because these functions are useful for other
purposes, they have been moved to a separate package called \texttt{fitdistrplus}.
Both the \texttt{mc2d} and the \texttt{fitdistrplus} packages are
available at the URL \url{https://r-forge.r-project.org/projects/riskassessment/}.

@@

According to international recommendations, a QRA should reflect the
{}``variability'' in the risk and calculate the {}``uncertainty''
associated with the risk estimate. The {}``variability'' represents
temporal, geographical and/or individual heterogeneity of the risk
for a given population. The {}``uncertainty'' is understood as stemming
from a lack of perfect knowledge about the QRA model structure and
associated parameters%
\footnote{In the engineering risk community, these concepts are refered as \textquotedbl{}aleatoric
uncertainty\textquotedbl{} for \textquotedbl{}variability\textquotedbl{}
and \textquotedbl{}epistemic uncertainty\textquotedbl{} for \textquotedbl{}uncertainty\textquotedbl{}.%
}.

In order to estimate the natural {}``variability'' of the risk,
a Monte-Carlo simulation approach may be useful: the empirical distribution
of the risk within the population may be estimated from the mathematical
combination of distributions reflecting the variability of parameters
across the population.

A two-dimensional (or second-order) Monte-Carlo simulation was proposed
to estimate the {}``uncertainty'' in the risk estimates stemming
from parameter uncertainty \cite{CULLEN-FREY-1999}. A two-dimensional
Monte-Carlo simulation is a Monte-Carlo simulation where the distributions
reflecting \textquotedbl{}variability\textquotedbl{} and the distributions
representing \textquotedbl{}uncertainty\textquotedbl{} are sampled
separately in the simulation, so that \textquotedbl{}variability\textquotedbl{}
and \textquotedbl{}uncertainty\textquotedbl{} in the output may be
estinated separately. It may be described as following (see Figure
\ref{fig:Mc2d}): 
\begin{enumerate}
\item The parameters of the model should be divided into three categories:
the parameters whose distributions reflect {}``variability only'',
hereinafter denoted as {}``variable parameters'', the parameters
whose distributions reflect {}``uncertainty only'', denoted as {}``uncertain
parameters'' and the parameters whose distributions reflect both
uncertainty and variability. For this latter category, a hierarchical
structure, using {}``hyper-parameters'', should be specified: if
a parameter is both uncertain and variable, one should be able to
specify an empirical or parametric distribution representing variability.
This distribution is conditional upon other parameters for which there
is some associated uncertainty. As an example, one should be able
to specify a relationship such as $X\left|a,b\right.\sim N\left(a,b\right)$,
where the specified normal distribution represents variability in
$x$ conditional upon parameters $a$ and $b$. Hyperdistributions,
such as $a\sim Unif\left(l_{a},u_{a}\right)$and $b\sim Unif\left(l_{b},u_{b}\right)$,
represent the uncertainty in the parameters $a$ and $b$;
\item A set of uncertain parameters are randomly sampled from their respective
distributions; 
\item The QRA is performed using a classical (one-dimensional) Monte-Carlo
simulation of size $N_{v}$, treating the uncertain parameters as
fixed. This QRA takes into account the variability in all variable
parameters, and leads to an empirical density function reflecting
the variability of exposure/risk across the population, conditional
upon the uncertain parameters. Various statistics (\emph{e.g.} the
mean, the standard deviation, percentiles) of the resulting empirical
density function are evaluated and stored; 
\item Steps 2) and 3) are performed a large number ($N_{u}$) of times ,
leading to $N_{u}$ sets of statistics; 
\item As output, the $50^{\text{th}}$ percentile (median) of each statistic
is used as a point estimate of this statistic; the $2.5^{\text{th}}$
and $97.5^{\text{th}}$ percentiles of each statistic are used to
establish a 95\% credible interval (CI95) of this statistic. The median
of the $N_{u}$ estimated values for each of the 101 estimated percentiles
allows us to display a {}``variability cumulative distribution''
via a graph. This curve is surrounded by the 2.5th and 97.5th percentiles
obtained from the $N_{u}$ estimates of each of the 101 percentiles.
\end{enumerate}
%
\begin{figure}
\noindent \begin{centering}
\includegraphics[scale=0.5,bb = 0 0 200 100, draft, type=eps]{mc2dsaumon}
\par\end{centering}

@@

\textquotedbl{}\texttt{mc2d}\textquotedbl{} is a set of R functions
that will help to develop such two-dimensional Monte-Carlo simulations.
The main difference from the procedure described above is that \texttt{mc2d}
uses arrays of (at least) two dimensions to derive the results: the
first dimension will reflect variability, the second will reflect
uncertainty. This document will not develop the method further, but
will illustrate the practical application of \texttt{mc2d}, using
a fictitious example.

@@

\textbf{Quantitative Risk Assessment:} \textbf{\emph{Escherichia coli}}
\textbf{O157:H7 infection linked to the consumption of frozen ground
beef in <3 year old children.}
\begin{itemize}
\item We assume that, in a given batch of ground beef, \emph{E. coli} O157:H7
are randomly distributed with a mean concentration of $c=10$ bacteria
(cfu) per gram of product;
\item We assume that no bacterial growth occurs in storage, since the product
is kept frozen until it is cooked, just before consumption;
\item 2.7\% of consumers cook their beef {}``rare'', 37.3\% {}``medium''
and 60.0\% {}``well done''; 
\item The following bacterial inactivation $i$ is associated with these
cooking practices:

\begin{itemize}
\item No inactivation for {}``rare'' cooking;
\item 1/5 surviving bacteria for a {}``medium'' cooking;
\item 1/50 surviving bacteria for a {}``well done'' cooking.
\end{itemize}
\item The variability in steak serving sizes $s$ for <3 year children was
estimated in a consumption survey. The {}``best fit'' to the data
was a gamma distribution with parameters: $shape=3.93,\: rate=0.0806$.
\item The dose-response relationship, describing the probability of illness,
$P,$ according to the dose is a one-hit model. The probability of
illness per hit $r$ is assumed to be constant with $r=0.001$.
\end{itemize}
The question is: {}``\emph{What is the distribution of the risk of
illness in the population that consumed the contaminated lot?}''

This distribution will be estimated using Monte-Carlo simulations
performed with R via the {}``\texttt{mc2d}'' package. First, the
model will be developed in a one dimensional framework. Then, in order
to include some uncertainties in the model, it will be derived in
a two dimensional framework.

@@

As a first step, we assume that no uncertainty exists in our model.
All distributions represent variability only. The model may be written
as:\begin{eqnarray*}
c & = & 10.\\
i & \sim & emp\left(\left\{ 1,1/5,1/50\right\} ,\left\{ 0.027,0.373,0.600\right\} \right)\\
s & \sim & gamma\left(3.93,0.0806\right)\\
n & \sim & Poisson\left(c\times i\times s\right)\\
P & = & 1-\left(1-0.001\right)^{n}\end{eqnarray*}

where $emp\left(\boldsymbol{X},\boldsymbol{P}\right)$ is an empirical
distribution wherein each value $X_{i}$ is associated with a probability
$P_{i}$.We will use a \textquotedbl{}classical\textquotedbl{} one
dimensional Monte-Carlo simulation, with 1,000 iterations. Using the
{}``\texttt{mc2d}'' package, the model may be written as:

@@

<<sh1, echo=true>>=
library(mc2d)
ndvar(1000)
conc <- 10
cook <- mcstoc(rempiricalD, values=c(1,1/5,1/50), prob=c(0.027,0.373,0.600))
serving <- mcstoc(rgamma,shape=3.93,rate=0.0806)
expo <- conc * cook * serving
dose <- mcstoc(rpois,lambda=expo)
r <- 0.001
risk <- 1-(1-r)^dose
EC1 <- mc(cook,serving,expo,dose,risk)
print(EC1)
summary(EC1)
@

@@

We will use a two-dimensional Monte-Carlo simulation, with 1,000 iterations
in the variability dimension and 100 iterations in the uncertainty
dimension. Using the {}``\texttt{mc2d}'' package, the model may
be written as:

<<sh2, echo=true>>=
ndunc(100)
conc <- mcstoc(rnorm,type="U",mean=10,sd=2)
cook <- mcstoc(rempiricalD, type="V",values=c(1,1/5,1/50), prob=c(0.027,0.373,0.600))
serving <- mcstoc(rgamma,type="V",shape=3.93,rate=0.0806)
expo <- conc * cook * serving
dose <- mcstoc(rpois,type="VU",lambda=expo)
r <- mcstoc(runif,type="U",min=0.0005,max=0.0015)
risk <- 1-(1-r)^dose
EC2 <- mc(conc,cook,serving,expo,dose,r,risk)
print(EC2,digits=2)
summary(EC2)
@

Note that the syntax is similar to the earlier model. However, a {}``\texttt{type}''
argument is provided for each distribution, indicating whether the
parameter distribution represents uncertainty (\texttt{type}=\textquotedbl{}\texttt{U}\textquotedbl{}),
variability (\texttt{type}=\textquotedbl{}V\textquotedbl{}), or both
(\texttt{type}=\textquotedbl{}\texttt{VU}\textquotedbl{}).

The summary provides estimates of the variability distributions (in
rows) but with a measure of their uncertainty, linked to the uncertainty
around \texttt{conc} and \texttt{r}. The estimate of the mean risk
is now uncertain. The median of the 100 simulations leads to a \textquotedbl{}best
estimate\textquotedbl{}of 0.0445, with a 95\% \textquotedbl{}credible
interval\textquotedbl{} of {[}0.191, 0.0730{]}.

@@

A typical session of R using {}``\texttt{mc2d}'' is as follows:
\begin{itemize}
\item From data, expert knowledge, \emph{etc.} an empirical or parametric
distribution is chosen for each {}``parent'' parameter. For developing
an empirical distribution from data , the {}``\texttt{fitdistrplus}''
package is recommended;
\item For each parameter, an \texttt{mcnode} object is constructed (key
functions: \texttt{mcdata, mcstoc}); 
\item Various \texttt{mcnode} objects are grouped into an \texttt{mc} object
(key function: \texttt{mc}).
\item The \texttt{mc} object is studied through summaries, graphs, and sensitivity
analysis (key functions: \texttt{summary.mc}, \texttt{plot.mc}, \texttt{tornado},
\texttt{tornadounc}).
\end{itemize}

@@

The {}``\texttt{mc2d}'' library should be loaded at the beginning
of your R session (\textquotedbl{}\texttt{library(mc2d)}\textquotedbl{}).

@@

An \texttt{mcnode} object is the basic element of an \texttt{mc} object.
It is an array of dimension $\left(nsv\times nsu\times nvariates\right)$
where $nsv$ is the dimension of variability, $nsu$ is the dimension
of uncertainty and $nvariates$ is the number of variates of the \texttt{mcnode}%
\footnote{In this section, we will only consider \texttt{mcnode}s with $nvariates=1$.%
}. Four types of \texttt{mcnode} exist: 
\begin{itemize}
\item \textquotedbl{}\texttt{V}\textquotedbl{} \texttt{mcnode}, for \textquotedbl{}\emph{Variability}\textquotedbl{},
is an array of dimension $\left(nsv\times1\times nvariates\right)$.
The distribution represents variability in the parameter; 
\item \textquotedbl{}\texttt{U}\textquotedbl{} \texttt{mcnode}, for \textquotedbl{}\emph{Uncertainty}\textquotedbl{},
is an array of dimension $\left(1\times nsu\times nvariates\right)$.
The distribution represents uncertainty in the parameter. 
\item \textquotedbl{}\texttt{VU}\textquotedbl{} \texttt{mcnode}, for \textquotedbl{}\emph{Variability
and Uncertainty}\textquotedbl{}, is an array of dimension $\left(nsv\times nsu\times nvariates\right)$.
The distribution represents both variability (in the first dimension)
and uncertainty (in the second dimension) in the parameter. 
\item Additionally, a \textquotedbl{}\texttt{0}\textquotedbl{} \texttt{mcnode}
is also defined . \textquotedbl{}\texttt{0}\textquotedbl{} stands
for \textquotedbl{}Neither Variability or Uncertainty\textquotedbl{}.
Such nodes are arrays of dimension $\left(1\times1\times nvariates\right)$.
No uncertainty or variability is considered for these nodes. A \textquotedbl{}\texttt{0}\textquotedbl{}
\texttt{mcnode} is not necessary in the univariate context (use a
scalar instead) but is useful in constructing multivariate nodes (See
section \ref{sec:Multivar}). 
\end{itemize}
%
\begin{figure}
\noindent \begin{centering}
\includegraphics[scale=0.4,bb = 0 0 200 100, draft, type=eps]{Illustration}
\par\end{centering}

\caption{\label{fig:Structure}Structure of the various \texttt{mcnode} objects.}

There are 5 ways to construct an \texttt{mcnode} object:
\begin{enumerate}
\item The \texttt{mcstoc} function constructs an \texttt{mcnode} from random
number generating functions;
\item The \texttt{mcdata} function constructs an \texttt{mcnode} from data
sets; 
\item \texttt{An mcnode} can be constructed directly from operations on
\texttt{mcnode} objects;
\item \texttt{mcprobtree} is a special function that constructs an \texttt{mcnode}
from other \texttt{mcnodes} using a probability tree;
\item Some functions, such as {}``\texttt{==}'' or {}``\texttt{>}''
, \texttt{is.na}, \texttt{is.finite} generate a new \texttt{mcnode}
when applied to an existing \texttt{mcnode}.
\end{enumerate}

@@

\begin{singlespace}
\noindent \begin{flushleft}
\texttt{mcstoc(func=runif, type=c(\textquotedbl{}V\textquotedbl{},
\textquotedbl{}U\textquotedbl{}, \textquotedbl{}VU\textquotedbl{},
\textquotedbl{}0\textquotedbl{}), ..., nsv=ndvar(), nsu=ndunc(), nvariates=1,
outm=\textquotedbl{}each\textquotedbl{}, nsample=\textquotedbl{}n\textquotedbl{},
seed=NULL, rtrunc=FALSE, linf=-Inf, lsup=Inf, lhs=FALSE)}
\par\end{flushleft}
\end{singlespace}
\begin{itemize}
\item \texttt{func} is a function providing random data or its name as a
character. The table \ref{tab:Distributions} provides available distributions
from the \texttt{stats} and the \texttt{mc2d} libraries that can be
used in \texttt{mcstoc}; 
\item \texttt{type} is the type of requested \texttt{mcnode}. By default,
\texttt{mcstoc} constructs a {}``\texttt{V}'' \texttt{mcnode};
\item \ldots{} are the arguments to be passed to the function \texttt{func},
with the exception of the argument providing the size of the sample.
This latter is calculated by the function according to \texttt{func},
\texttt{type}, \texttt{nsv,} \texttt{nsu} and \texttt{nvariates}.
If the name of the argument specifying the size of the sample is not
\texttt{n} (\emph{e.g.} functions \texttt{rhyper} and \texttt{rwilcox},
see table \ref{tab:Distributions}), the name of this parameter should
be provided in the \texttt{nsample} argument. \emph{Note that all
of the following arguments should be named};
\item \texttt{nsv} and \texttt{nsu} are the number of samples needed in
the variability and uncertainty dimension, respectively. By default,
these values are the ones provided by \texttt{ndvar()} and \texttt{ndunc()},
respectively;
\item \texttt{nvariates} is the desired number of variates in the \texttt{mcnode};
\item \texttt{outm} is the default output for multivariate nodes;
\item \texttt{seed} optionally specifies a seed for the random number generator;
\item \texttt{rtrunc} allows truncation of a distribution between \texttt{linf}
and \texttt{lsup}. This function is not valid for every distribution
(see table \ref{tab:Distributions}). See the \texttt{rtrunc} function
help for further details;
\item \texttt{lhs} allows Latin hypercube sampling of the node . This function
is not valid for every distribution (see table \ref{tab:Distributions}).
See the \texttt{lhs} function help for further details. 
\end{itemize}
%
\begin{table}
\caption{\label{tab:Distributions}Available distributions}

@@

<<sh3, echo=true, eval=FALSE>>=
conc <- mcstoc(rnorm,type="U",mean=10,sd=2)
cook <- mcstoc(rempiricalD, type="V",values=c(1,1/5,1/50), prob=c(0.027,0.373,0.600))
serving <- mcstoc(rgamma,type="V",shape=3.93,rate=0.0806)
...
dose <- mcstoc(rpois,type="VU",lambda=expo)
r <- mcstoc(runif,type="U",min=0.0005,max=0.0015)
...
@

@@

<<sh3bis, echo=true, eval=true>>=
x <- mcstoc(rnorm, mean=2, sd=3, rtrunc=TRUE, linf=1.5, lsup=2, lhs=TRUE)
summary(x)
@

@@

The functions {}``\texttt{==}'', ''\texttt{<}'', ''\texttt{<}='',
''\texttt{>=}'', {}``\texttt{>}'', generate an \texttt{mcnode}
when applied to another \texttt{mcnode}.

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%% mc2d: Tools for Two-Dimensional Monte-Carlo Simulations
%%%
%%%         Sweave vignette file
%%%

\documentclass[10pt, english]{article}

%\usepackage[letterpaper]{geometry}
\usepackage[T1]{fontenc}
\usepackage[latin9]{inputenc}
\usepackage{amsmath}
\usepackage{graphicx}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% User specified LaTeX commands.
% sweave commands for vignette
%\VignetteIndexEntry{Tools for Two-Dimensional Monte-Carlo Simulations}
%\VignettePackage{mc2d}
%\VignetteDepends{mvtnorm}

\title{The \texttt{mc2d} package.}
\author{R. POUILLOT, M.-L. DELIGNETTE-MULLER, D.L. KELLY \& J.-B. DENIS}

\begin{document}
\maketitle

This documentation is intended for readers with:
\begin{itemize}
\item A medium level of experience in R. Please refer to the Manual An
Introduction to R available with R distribution if needed;
\item Some knowledge about Monte-Carlo simulation (its basic principles
and its utility) and about Quantitative Risk Assessment (QRA). 
\end{itemize}

This documentation will not describe all arguments of the functions.
The definitive reference remains the documentation associated with
the package.

\tableofcontents{}

\section{Introduction}
\subsection{What is \texttt{mc2d}?}

\texttt{mc2d} means Two-Dimensional Monte-Carlo (\emph{Monte-Carlo
à Deux Dimensions}). This package :
\begin{itemize}
\item provides additional probability distributions;
\item provides tools to construct One-Dimensional and Two-Dimensional Monte-Carlo
Simulations;
\item provides tools to analyse One-Dimensional and Two-Dimensional Monte-Carlo
Simulations.
\end{itemize}

In a previous version, some tools to fit parametric distributions
to data were included. Because these functions are useful for other
purposes, they have been moved to a separate package called \texttt{fitdistrplus}.

@@

According to international recommendations, a QRA should reflect the
variability in the risk and calculate the uncertainty
associated with the risk estimate. The variability represents
temporal, geographical and/or individual heterogeneity of the risk
for a given population. The uncertainty is understood as stemming
from a lack of perfect knowledge about the QRA model structure and
associated parameters%
\footnote{In the engineering risk community, these concepts are refered as aleatoric
uncertainty for variability
and epistemic uncertainty for uncertainty.%
}.

In order to estimate the natural variability of the risk,
a Monte-Carlo simulation approach may be useful: the empirical distribution
of the risk within the population may be estimated from the mathematical
combination of distributions reflecting the variability of parameters
across the population.

A two-dimensional (or second-order) Monte-Carlo simulation was proposed
to estimate the uncertainty in the risk estimates stemming
from parameter uncertainty \cite{CULLEN-FREY-1999}. A two-dimensional
Monte-Carlo simulation is a Monte-Carlo simulation where the distributions
reflecting variability and the distributions
representing uncertainty are sampled
separately in the simulation, so that variability
and uncertainty in the output may be
estinated separately. It may be described as following (see Figure
\ref{fig:Mc2d}):

\begin{enumerate}
\item The parameters of the model should be divided into three categories:
the parameters whose distributions reflect variability only,
hereinafter denoted as variable parameters, the parameters
whose distributions reflect uncertainty only, denoted as uncertain
parameters and the parameters whose distributions reflect both
uncertainty and variability. For this latter category, a hierarchical
structure, using hyper-parameters, should be specified: if
a parameter is both uncertain and variable, one should be able to
specify an empirical or parametric distribution representing variability.
This distribution is conditional upon other parameters for which there
is some associated uncertainty. As an example, one should be able
to specify a relationship such as
\[ X\left|a,b\right.\sim N\left(a,b\right) \]
where the specified normal distribution represents variability in
$x$ conditional upon parameters $a$ and $b$. Hyperdistributions,
such as
\[ a\sim Unif\left(l_{a},u_{a}\right) \]
and
\[ b\sim Unif\left(l_{b},u_{b}\right) \]
represent the uncertainty in the parameters $a$ and $b$;
\item A set of uncertain parameters are randomly sampled from their respective
distributions; 
\item The QRA is performed using a classical (one-dimensional) Monte-Carlo
simulation of size $N_{v}$, treating the uncertain parameters as
fixed. This QRA takes into account the variability in all variable
parameters, and leads to an empirical density function reflecting
the variability of exposure/risk across the population, conditional
upon the uncertain parameters. Various statistics (\emph{e.g.} the
mean, the standard deviation, percentiles) of the resulting empirical
density function are evaluated and stored; 
\item Steps 2) and 3) are performed a large number ($N_{u}$) of times ,
leading to $N_{u}$ sets of statistics; 
\item As output, the $50^{th}$ percentile (median) of each statistic
is used as a point estimate of this statistic; the $2.5^{th}$
and $97.5^{th}$ percentiles of each statistic are used to
establish a 95\% credible interval (CI95) of this statistic. The median
of the $N_{u}$ estimated values for each of the 101 estimated percentiles
allows us to display a variability cumulative distribution
via a graph. This curve is surrounded by the 2.5th and 97.5th percentiles
obtained from the $N_{u}$ estimates of each of the 101 percentiles.
\end{enumerate}
%
\begin{figure}
\noindent \begin{centering}
\includegraphics[angle=270, width=1\textwidth]{mc2dsaumon}
\par\end{centering}

@@

\texttt{mc2d} is a set of R functions
that will help to develop such two-dimensional Monte-Carlo simulations.
The main difference from the procedure described above is that \texttt{mc2d}
uses arrays of (at least) two dimensions to derive the results: the
first dimension will reflect variability, the second will reflect
uncertainty. This document will not develop the method further, but
will illustrate the practical application of \texttt{mc2d}, using
a fictitious example.

@@

\textbf{Quantitative Risk Assessment: \emph{Escherichia coli}
O157:H7 infection linked to the consumption of frozen ground
beef in <3 year old children.}
\begin{itemize}
\item We assume that, in a given batch of ground beef, \emph{E. coli} O157:H7
are randomly distributed with a mean concentration of $c=10$ bacteria
(cfu) per gram of product;
\item We assume that no bacterial growth occurs in storage, since the product
is kept frozen until it is cooked, just before consumption;
\item 2.7\% of consumers cook their beef rare, 37.3\% medium
and 60.0\% well done; 
\item The following bacterial inactivation $i$ is associated with these
cooking practices:

\begin{itemize}
\item No inactivation for rare cooking;
\item 1/5 surviving bacteria for a medium cooking;
\item 1/50 surviving bacteria for a well done cooking.
\end{itemize}
\item The variability in steak serving sizes $s$ for <3 year children was
estimated in a consumption survey. The best fit to the data
was a gamma distribution with parameters: $shape=3.93,\: rate=0.0806$.
\item The dose-response relationship, describing the probability of illness,
$P,$ according to the dose is a one-hit model. The probability of
illness per hit $r$ is assumed to be constant with $r=0.001$.
\end{itemize}
The question is: \emph{What is the distribution of the risk of
illness in the population that consumed the contaminated lot?}

This distribution will be estimated using Monte-Carlo simulations
performed with R via the \texttt{mc2d} package. First, the
model will be developed in a one dimensional framework. Then, in order
to include some uncertainties in the model, it will be derived in
a two dimensional framework.

@@

As a first step, we assume that no uncertainty exists in our model.
All distributions represent variability only. The model may be written
as:
\begin{eqnarray*}
c & = & 10.\\
i & \sim & emp\left(\left\{ 1,1/5,1/50\right\} ,\left\{ 0.027,0.373,0.600\right\} \right)\\
s & \sim & gamma\left(3.93,0.0806\right)\\
n & \sim & Poisson\left(c\times i\times s\right)\\
P & = & 1-\left(1-0.001\right)^{n}
\end{eqnarray*}

where $emp\left(X,P\right)$ is an empirical
distribution wherein each value $X_{i}$ is associated with a probability
$P_{i}$.We will use a classical one
dimensional Monte-Carlo simulation, with 1,000 iterations. Using the
\texttt{mc2d} package, the model may be written as:

@@

%true
<<sh1, eval=true, echo=true>>=
library(mc2d)
ndvar(1000)
conc <- 10
cook <- mcstoc(rempiricalD, values=c(1,1/5,1/50), prob=c(0.027,0.373,0.600))
serving <- mcstoc(rgamma,shape=3.93,rate=0.0806)
expo <- conc * cook * serving
dose <- mcstoc(rpois,lambda=expo)
r <- 0.001
risk <- 1-(1-r)^dose
EC1 <- mc(cook,serving,expo,dose,risk)
print(EC1)
summary(EC1)
@

@@

We will use a two-dimensional Monte-Carlo simulation, with 1,000 iterations
in the variability dimension and 100 iterations in the uncertainty
dimension. Using the \texttt{mc2d} package, the model may
be written as:

%true
<<sh2, eval=true, echo=true>>=
ndunc(100)
conc <- mcstoc(rnorm,type="U",mean=10,sd=2)
cook <- mcstoc(rempiricalD, type="V",values=c(1,1/5,1/50), prob=c(0.027,0.373,0.600))
serving <- mcstoc(rgamma,type="V",shape=3.93,rate=0.0806)
expo <- conc * cook * serving
dose <- mcstoc(rpois,type="VU",lambda=expo)
r <- mcstoc(runif,type="U",min=0.0005,max=0.0015)
risk <- 1-(1-r)^dose
EC2 <- mc(conc,cook,serving,expo,dose,r,risk)
print(EC2,digits=2)
summary(EC2)
@

Note that the syntax is similar to the earlier model. However, a \texttt{type}
argument is provided for each distribution, indicating whether the
parameter distribution represents uncertainty (\texttt{type}=\texttt{U}),
variability (\texttt{type}=V), or both
(\texttt{type}=\texttt{VU}).

The summary provides estimates of the variability distributions (in
rows) but with a measure of their uncertainty, linked to the uncertainty
around \texttt{conc} and \texttt{r}. The estimate of the mean risk
is now uncertain. The median of the 100 simulations leads to a best
estimateof 0.0445, with a 95\% credible
interval of {[}0.191, 0.0730{]}.

@@

A typical session of R using \texttt{mc2d} is as follows:
\begin{itemize}
\item From data, expert knowledge, \emph{etc.} an empirical or parametric
distribution is chosen for each parent parameter. For developing
an empirical distribution from data , the \texttt{fitdistrplus}
package is recommended;
\item For each parameter, an \texttt{mcnode} object is constructed (key
functions: \texttt{mcdata, mcstoc}); 
\item Various \texttt{mcnode} objects are grouped into an \texttt{mc} object
(key function: \texttt{mc}).
\item The \texttt{mc} object is studied through summaries, graphs, and sensitivity
analysis (key functions: \texttt{summary.mc}, \texttt{plot.mc}, \texttt{tornado},
\texttt{tornadounc}).
\end{itemize}

@@

The \texttt{mc2d} library should be loaded at the beginning
of your R session (\texttt{library(mc2d)}).

@@

An \texttt{mcnode} object is the basic element of an \texttt{mc} object.
It is an array of dimension \texttt{nsv} * \texttt{nsu} * \texttt{nvariates}
where $nsv$ is the dimension of variability, $nsu$ is the dimension
of uncertainty and $nvariates$ is the number of variates of the \texttt{mcnode}%
\footnote{In this section, we will only consider \texttt{mcnode}s with $nvariates=1$.%
}. Four types of \texttt{mcnode} exist: 
\begin{itemize}
\item \texttt{V} \texttt{mcnode}, for \emph{Variability},
is an array of dimension $\left(nsv\times1\times nvariates\right)$.
The distribution represents variability in the parameter; 
\item \texttt{U} \texttt{mcnode}, for \emph{Uncertainty},
is an array of dimension $\left(1\times nsu\times nvariates\right)$.
The distribution represents uncertainty in the parameter. 
\item \texttt{VU} \texttt{mcnode}, for \emph{Variability
and Uncertainty}, is an array of dimension $\left(nsv\times nsu\times nvariates\right)$.
The distribution represents both variability (in the first dimension)
and uncertainty (in the second dimension) in the parameter. 
\item Additionally, a \texttt{0} \texttt{mcnode}
is also defined . \texttt{0} stands
for Neither Variability or Uncertainty.
Such nodes are arrays of dimension $\left(1\times1\times nvariates\right)$.
No uncertainty or variability is considered for these nodes. A \texttt{0}
\texttt{mcnode} is not necessary in the univariate context (use a
scalar instead) but is useful in constructing multivariate nodes (See
section \ref{sec:Multivar}). 
\end{itemize}
%
\begin{figure}
\noindent \begin{centering}
\includegraphics[angle=270, width=.5\textwidth]{Illustration}
\par\end{centering}

\caption{\label{fig:Structure}Structure of the various \texttt{mcnode} objects.}
\end{figure}

There are 5 ways to construct an \texttt{mcnode} object:
\begin{enumerate}
\item The \texttt{mcstoc} function constructs an \texttt{mcnode} from random
number generating functions;
\item The \texttt{mcdata} function constructs an \texttt{mcnode} from data
sets; 
\item \texttt{An mcnode} can be constructed directly from operations on
\texttt{mcnode} objects;
\item \texttt{mcprobtree} is a special function that constructs an \texttt{mcnode}
from other \texttt{mcnodes} using a probability tree;
\item Some functions, such as \texttt{==} or \texttt{>}
, \texttt{is.na}, \texttt{is.finite} generate a new \texttt{mcnode}
when applied to an existing \texttt{mcnode}.
\end{enumerate}

@@

%\begin{singlespace}
\noindent \begin{flushleft}
\texttt{mcstoc(func=runif, type=c(V,
U, VU,
0), ..., nsv=ndvar(), nsu=ndunc(), nvariates=1,
outm=each, nsample=n,
seed=NULL, rtrunc=FALSE, linf=-Inf, lsup=Inf, lhs=FALSE)}
\par\end{flushleft}
%\end{singlespace}
\begin{itemize}
\item \texttt{func} is a function providing random data or its name as a
character. The table \ref{tab:Distributions} provides available distributions
from the \texttt{stats} and the \texttt{mc2d} libraries that can be
used in \texttt{mcstoc}; 
\item \texttt{type} is the type of requested \texttt{mcnode}. By default,
\texttt{mcstoc} constructs a \texttt{V} \texttt{mcnode};
\item \ldots{} are the arguments to be passed to the function \texttt{func},
with the exception of the argument providing the size of the sample.
This latter is calculated by the function according to \texttt{func},
\texttt{type}, \texttt{nsv,} \texttt{nsu} and \texttt{nvariates}.
If the name of the argument specifying the size of the sample is not
\texttt{n} (\emph{e.g.} functions \texttt{rhyper} and \texttt{rwilcox},
see table \ref{tab:Distributions}), the name of this parameter should
be provided in the \texttt{nsample} argument. \emph{Note that all
of the following arguments should be named};
\item \texttt{nsv} and \texttt{nsu} are the number of samples needed in
the variability and uncertainty dimension, respectively. By default,
these values are the ones provided by \texttt{ndvar()} and \texttt{ndunc()},
respectively;
\item \texttt{nvariates} is the desired number of variates in the \texttt{mcnode};
\item \texttt{outm} is the default output for multivariate nodes;
\item \texttt{seed} optionally specifies a seed for the random number generator;
\item \texttt{rtrunc} allows truncation of a distribution between \texttt{linf}
and \texttt{lsup}. This function is not valid for every distribution
(see table \ref{tab:Distributions}). See the \texttt{rtrunc} function
help for further details;
\item \texttt{lhs} allows Latin hypercube sampling of the node . This function
is not valid for every distribution (see table \ref{tab:Distributions}).
See the \texttt{lhs} function help for further details. 
\end{itemize}
%
\begin{table}
\caption{\label{tab:Distributions}Available distributions}

@@

%false
<<sh3, echo=true, eval=FALSE>>=
conc <- mcstoc(rnorm,type="U",mean=10,sd=2)
cook <- mcstoc(rempiricalD, type="V",values=c(1,1/5,1/50), prob=c(0.027,0.373,0.600))
serving <- mcstoc(rgamma,type="V",shape=3.93,rate=0.0806)
...
dose <- mcstoc(rpois,type="VU",lambda=expo)
r <- mcstoc(runif,type="U",min=0.0005,max=0.0015)
...
@

@@

%true
<<sh3bis, echo=true, eval=true>>=
x <- mcstoc(rnorm, mean=2, sd=3, rtrunc=TRUE, linf=1.5, lsup=2, lhs=TRUE)
summary(x)
@

@@

The functions \texttt{==}, \texttt{<}, \texttt{<}=,
\texttt{>=}, \texttt{>}, generate an \texttt{mcnode}
when applied to another \texttt{mcnode}.

Special functions \texttt{is.na(x)}, \texttt{is.nan(x)}, \texttt{is.finite(x)},
\texttt{is.infinite(x)} are implemented to test if any values are
\texttt{NA} (missing data), \texttt{NaN} (\emph{Not A Number}),
or finite .
